In the Age of Web:\\ \textnormal{\LARGE Typed Functional-First Programming Revisited}

### 2015-05-01 ml-2014-paper/paper.tex
Fix typos (ML 2014 post)

\begin{abstract}
% What is the problem
Most programming languages were designed before the age of web.
% Why is this a problem
This matters because the web changes many assumptions that typed functional language designers take 
for granted. For example, programs do not run in a closed word, but must instead interact with 
(changing and likely unreliable) services and data sources, communication is often asynchronous 
or event-driven, and programs need to interoperate with untyped environments.

% What is my solution
In this paper, we present how F\# language and libraries face the challenges posed by the web.
Technically, this comprises using \emph{type providers} for integration with external information 
sources and for integration with untyped programming environments, using \emph{lightweight 
meta-programming} for targeting JavaScript and \emph{computation expressions} for writing 
asynchronous code.

% Why does the solution matter
In this inquiry, the holistic perspective is more important than each of the features in isolation.
We use a practical case study as a starting point and look how F\# language and libraries approach 
the challenges posed by the web. The specific lessons learned are perhaps less interesting than 
our attempt to uncover hidden assumptions that no longer hold in the age of web.
\end{abstract}

@@

We do not claim that the F\# approach is the only possible. Rather, this paper should be seen
as a programming language experiment \cite{philosophy-pl} or an empirical observation of the 
approach used by the F\# community. We aim to provide an intriguing exploration of hidden assumptions 
and present what can be achieved using a combination of F\# features. We do so by starting with 
a simple, yet real-world problem and then exploring a solution. More specifically, the contributions 
of this position paper are:

\begin{itemize}
\item We present a case study (Section~\ref{sec:case}) showing how a combination of numerous
  F\# language features can be used for the development of modern web applications. This
  is not a toy demonstration, but an example of how F\# is used in the industry.

@@

\item Throughout the paper, we discuss how the age of the web breaks the assumptions commonly 
  taken as granted in typed functional programming. We revisit the notion of type safety
  in the context of web (Section~\ref{sec:tp-relative}) and the notion of fixed language 
  semantics (Section~\ref{sec:js-relative}).
\end{itemize}

@@

The main part of the sample program is a function \ident{render} that asynchronously fetches 
data for selected countries and generates a chart. To keep the code simple, we iterate over the 
\ident{infos} list from the previous section and load data for countries one by one:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{render}~()~=~\ident{async}~\{ \\
 \quad \kvd{let}~\ident{head}~=~\str{School enrollment, tertiary (\% gross)} \\
 \quad \kvd{let}~\ident{o}~=~\ident{h.HighchartsOptions}() \\
 \quad \ident{o.chart} \leftarrow \ident{h.HighchartsChartOptions}(\ident{renderTo}=\str{plc}) \\
 \quad \ident{o.title} \leftarrow \ident{h.HighchartsTitleOptions}(\ident{text}~=~\ident{head}) \\
 \quad \ident{o.series} \leftarrow \lbrack| ~ |\rbrack
 \\[0.5em]   
 \quad \kvd{for}~\ident{name},~\ident{ind},~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do}\\
 \qquad   \kvd{if}~\ident{unbox}\langl \ident{bool} \rangl~(\ident{check.is}(\str{:checked}))~\kvd{then} \\
 \qquad\quad     \kvd{let!}~\ident{v}~=~\ident{ind.}\lident{School enrollment, tertiary (\% gross)} \\
 \qquad\quad     \kvd{let}~\ident{data}~=~\ident{vals}~
                   |\rang~\ident{Seq.map}~(\kvd{fun}~(\ident{k}, \ident{v})~\rightarrow \lbrack|~\ident{number~k};~\ident{number~v} ~|\rbrack)~
                   |\rang~\ident{Array.ofSeq} \\
 \qquad\quad     \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name})) ~\}
\end{array}
\end{equation*}
%
Although the function looks like ordinary code, it is wrapped in the $\ident{async}~\{\ldots\}$ 
block, which is an F\# computation expression \cite{fsharp-zoo}. The F\# compiler performs 
de-sugaring similar to the CPS transformation and interprets keywords such as \kvd{let!} and 
\kvd{for} using special operations (monadic bind and other). The \ident{async} identifier 
determines that we are writing asynchronous workflow \cite{fsharp-async} that makes it 
possible to include non-blocking calls in the block.

@@

Two notable points here are that \ident{async} translated to JavaScript is restricted to a single 
thread, which is not the case for ordinary F\# code (Section~\ref{sec:js-async}) and that the 
\ident{HighchartOptions} object preserves some of the underlying JavaScript semantics 
(Section~\ref{sec:js-lib}). Finally, the last part of the example code registers event handlers that 
redraw the chart when checkbox is clicked:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{for}~\ignp, \ignp,~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do} \\
 \quad \ident{check.click}(\kvd{fun}~\ignp~\rightarrow \ident{Async.StartImmediate}(\ident{render}()))
\end{array}
\end{equation*}
%
The \ident{click} operation (exposed by jQuery) takes a function that should be called when the 
event occurs. Calling it is a side-effectful operation that registers the handler. As \ident{render} 
is an asynchronous operation, we invoke it using the \ident{StartImmediate} primitive from the 
F\# library, which starts the computation without waiting for the result (the only way to start
a non-blocking operation in JavaScript).

@@

The aim of this paper is not to make value judgements and argue what is better. Using the case
study as a basis, we claim that the outlined approach is \emph{one possible} and that it
\emph{works in practice}. In the rest of the paper, we give more details about the most important 
aspects of the approach and discuss alternatives.

@@

Type providers \cite{fsharp-typeprov} is a mechanism for integrating external components into a 
statically typed programming language. Such components include information sources (such as World
Bank), other environments (here, JavaScript libraries via TypeScript), but type-providers can 
also be used for limited meta-programming (the \ident{Asynchronous} parameter can be seen as
a form of meta-programming).

@@

\begin{itemize}
\item \emph{Type provider failure.} 
  The projection (a type provider) is implemented in F\# and can fail (for example, if internet
  is required but unavailable). However, when it succeeds, it generates a valid F\# code and so 
  unchecked compilation errors cannot happen.

@@

The World Bank provider \cite{fsharp-data} is a type provider for a specific data source -- it has 
been designed specifically for World Bank data. This contrasts with F\# Data type providers for 
working with CSV, XML and JSON data that take sample or schema as parameter and can thus be used 
with files of any structure.

@@

\begin{itemize}
\item If you are offline when using the type provider (compiling or editing), the type provider will
  not be able to obtain the list of countries or indicators. This is an example of \emph{type provider
  failure}. In case of World Bank, this is partly prevented by caching of the schema, but internet
  access is required when using the provider for the first time.

@@

\noindent
When contrasted with the traditional view of safety in ML family of languages, the World Bank 
type provider certainly relaxes the usual safety conditions. However, this is not the case when 
compared with standard developer practices. In particular, if we wrote 
$\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$ by hand, the \emph{runtime failure}
behaviour would be the same, but we would lose the useful \emph{recompilation failure}. The following
section discusses this (and other) alternatives.

@@

\vspace{-1em}
\paragraph{Accessing countries without type providers.} First, consider how would we write the code to obtain a list
of countries without using the World Bank type provider. Assuming we want only the countries and
regions in Figure~\ref{fig:wb}, we could specify the country codes and then obtain the country objects:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countryCodes}~=~\lbrack~\str{EU},~\str{CZE},~\str{GBP},~\str{USA}~\rbrack \\
 \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry})
\end{array}
\end{equation*}
%
The motivation for the case study was a data journalism application with a transparent logic
that is easy to understand (and perhaps even modify). The type provider certainly achieves these
goals better -- a country can be added just by looking through an auto-complete list and we do not
need to know its code or exact name. However, if we wanted to list all countries in a region, then
the World Bank type provider would not be helpful.

@@

\vspace{-1em}
\paragraph{Providing safer data accessors.} 
The \emph{recompilation failure} when a country is removed from the dataset is an indication that
code would not behave as expected as runtime and so we find it useful. The problematic case is 
a \emph{runtime failure} when a country is removed after the program is compiled. One way to 
avoid this error would be to change the type of the provided members like
\ident{data.}\lident{Czech Republic} from \ident{Country} to \ident{Country~option}.

@@

Whether using option types is more desirable or not often depends on the use case (how we
want to handle errors) and reliability of the data source (countries do not disappear often).
This alternative can be supported by adding a static parameter to the type provider (like
$\ident{Asynchronous}=\kvd{true}$). For example, the CSV type provider from F\# Data has a 
parameter \ident{AssumeMissingValues} that serves exactly this purpose.

@@

For type providers that provide access to external information sources, we can formulate a 
\emph{relativized} notion of type safety. A full formalization is beyond the scope of this
paper, but the following provides an outline of such theorem.

@@

JavaScript has become the \emph{lingua franca} of the web and increasing number of programming 
languages provide a way of compiling to JavaScript. In F\#, the first project was F\# WebTools
\cite{fsharp-webtools} in 2007. A more recent and complex framework called Websharper \cite{websharper-guis} 
is available with full commercial support. In this paper, we use FunScript which is a lightweight
library focused just on translating F\# to JavaScript.

In this section, we discuss the options available when targeting JavaScript. The compilation to 
JavaScript can be implemented as compiler back-end or as a library based on meta-programming 
(Section~\ref{sec:js-meta}), there are different approaches to base libraries (Section~\ref{sec:js-lib}),
asynchronous computing (Section~\ref{sec:js-async}) and semantic mismatch between the host language
and JavaScript (Section~\ref{sec:js-relative}).

@@

In FunScript, we typically need to translate the entire source file. To allow this, we instruct the
compiler to store code as data for the whole module using the \ident{ReflectedDefinition} attribute.
This is done at the beginning of the file (which we omitted in Section~\ref{sec:case}):
%
\begin{equation*}
\begin{array}{l}
 \lbrack\langl\ident{ReflectedDefinition}\rangl\rbrack \\
 \kvd{module}~\ident{Program}
 \\[0.5em]
 \kvd{open}~\ident{FunScript} \\
 \kvd{open}~\ident{FunScript.TypeScript} \\
 \kvd{open}~\ident{FSharp.Data} 
\end{array}
\end{equation*}
%
When marked with \ident{ReflectedDefinition}, the F\# compiler stores the body of all functions and 
methods in the marked module as \emph{quotations} that can be retrieved at run-time. Note that this does
not store the AST for the entire file -- for example, information about declared types need to be 
obtained separately using the .NET reflection mechanism.

@@

Using the lightweight meta-programming approach in FunScript would not be appropriate if we wanted to 
compile arbitrary existing F\# source code to JavaScript without any modifications. For that, using a 
compiler back-end is a better choice. However, for the task solved in our case study (compiling 
newly written F\# code), the approach works well. A related question is providing mapping for 
standard (in our case, F\# and .NET) libraries used by the code. We discuss this in the next section.

@@

The F\# code in our case study uses a number of libraries. This includes the F\# core library (for 
example, the \ident{List.map} function), standard .NET libraries (iteration using the \kvd{for} loop
uses .NET \ident{IEnumerable} interface) and JavaScript libraries. Accessing these libraries in 
FunScript follows similar approach to F\# WebTools \cite{fsharp-webtools}.

@@

FunScript implements (explicit) mappings for standard F\# and .NET libraries that are useful in 
JavaScript (collections, string manipulation, date and other). Those are either reimplemented in
F\# with \ident{ReflectedDefinition} attribute, or mapped to standard JavaScript library.

@@

\noindent
In summary, the approach used by F\# is to reuse as much of F\# and .NET as possible. This
makes writing and reusing code easier and lets developers use existing familiar libraries and
idioms. However, it means that semantics of F\# running as JavaScript is not the same as the
semantics of F\# running as compiled code. We return to this topic in Section~\ref{sec:js-relative}. 
Before that, the following section gives one more example where similar approach is used.

@@

In F\# asynchronous workflows \cite{fsharp-async} serve two purposes. First, they provide a
way for running multiple tasks in parallel. Second, they make it possible to write long-running
non-blocking code without the use of explicit callbacks. In our case study, we only use the 
latter aspect. When loading data for selected countries in a loop, we fetch data asynchronously 
(using an AJAX call). Under the cover, JavaScript triggers a callback, which then resumes the 
asynchronous workflow.

@@

In ordinary F\#, asynchronous workflows can be started in a number of ways: \ident{Async.RunSynchronously}
starts the work and blocks until it completes; \ident{Async.Start} starts the work in the background
and \ident{Async.\-Start\-Immediate} starts the work on the current thread and runs callbacks on 
in the same synchronization context (essentially on the same thread). However, only the last one
can be mapped to JavaScript (FunScript translator will fail if the other two are used).

@@

There is a number of projects that solve similar problems to the ones described in this paper.
In the F\# ecosystem, the first project compiling to JavaScript was F\# WebTools \cite{fsharp-webtools},
which also supported asynchronous computations (although using a separate computation type)
and interop with JavaScript. More recently, WebSharper \cite{websharper-piglets,websharper-guis}
is a more complete framework for web development that also provide composable abstractions for
building forms (based on formlets \cite{links-formlets}) and entire applications.

@@

The key idea of this paper is that writing modern applications for the web requires us to 
reconsider many assumptions that functional language designers take for granted. The novelty
of this paper is not in any single technology it presents, but in the combination it shows.
For this reason, we used the form of a cast study, using a significant example to guide our
discussion.

>>>>

\begin{abstract}
% What is the problem
Most programming languages were designed before the age of web.
% Why is this a problem
This matters because the web changes many assumptions that typed functional language designers take
for granted. For example, programs do not run in a closed world, but must instead interact with
(changing and likely unreliable) services and data sources, communication is often asynchronous
or event-driven, and programs need to interoperate with untyped environments.

% What is my solution
In this paper, we present how the F\# language and libraries face the challenges posed by the web.
Technically, this comprises using \emph{type providers} for integration with external information
sources and for integration with untyped programming environments, using \emph{lightweight
meta-programming} for targeting JavaScript and \emph{computation expressions} for writing
asynchronous code.

% Why does the solution matter
In this inquiry, the holistic perspective is more important than each of the features in isolation.
We use a practical case study as a starting point and look at how F\# language and libraries
approach the challenges posed by the web. The specific lessons learned are perhaps less interesting
than our attempt to uncover hidden assumptions that no longer hold in the age of web.
\end{abstract}

@@

We do not claim that the F\# approach is the only possible one. Rather, this paper should be seen
as a programming language experiment \cite{philosophy-pl} or an empirical observation of the
approach used by the F\# community. We aim to provide an intriguing exploration of hidden assumptions
and present what can be achieved using a combination of F\# features. We do so by starting with
a simple, yet real-world problem and then exploring a solution. More specifically, the contributions
of this position paper are:

\begin{itemize}
\item We present a case study (Section~\ref{sec:case}) showing how a combination of numerous
  F\# language features can be used for the development of modern web applications. This
  is not a toy demonstration, but an example of how F\# is used in industry.

@@

\item Throughout the paper, we discuss how the age of the web breaks the assumptions commonly
  taken for granted in typed functional programming. We revisit the notion of type safety
  in the context of the web (Section~\ref{sec:tp-relative}) and the notion of fixed language
  semantics (Section~\ref{sec:js-relative}).
\end{itemize}

@@

The main part of the sample program is a function \ident{render} that asynchronously fetches
data for selected countries and generates a chart. To keep the code simple, we iterate over the
\ident{infos} list from the previous section and load data for countries one by one:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{render}~()~=~\ident{async}~\{ \\
 \quad \kvd{let}~\ident{head}~=~\str{School enrollment, tertiary (\% gross)} \\
 \quad \kvd{let}~\ident{o}~=~\ident{h.HighchartsOptions}() \\
 \quad \ident{o.chart} \leftarrow \ident{h.HighchartsChartOptions}(\ident{renderTo}=\str{plc}) \\
 \quad \ident{o.title} \leftarrow \ident{h.HighchartsTitleOptions}(\ident{text}~=~\ident{head}) \\
 \quad \ident{o.series} \leftarrow \lbrack| ~ |\rbrack
 \\[0.5em]
 \quad \kvd{for}~\ident{name},~\ident{ind},~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do}\\
 \qquad   \kvd{if}~\ident{unbox}\langl \ident{bool} \rangl~(\ident{check.is}(\str{:checked}))~\kvd{then} \\
 \qquad\quad     \kvd{let!}~\ident{v}~=~\ident{ind.}\lident{School enrollment, tertiary (\% gross)} \\
 \qquad\quad     \kvd{let}~\ident{data}~=~\ident{vals}~
                   |\rang~\ident{Seq.map}~(\kvd{fun}~(\ident{k}, \ident{v})~\rightarrow \lbrack|~\ident{number~k};~\ident{number~v} ~|\rbrack)~
                   |\rang~\ident{Array.ofSeq} \\
 \qquad\quad     \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name})) ~\}
\end{array}
\end{equation*}
%
Although the function looks like ordinary code, it is wrapped in the $\ident{async}~\{\ldots\}$
block, which is an F\# computation expression \cite{fsharp-zoo}. The F\# compiler performs
de-sugaring similar to the CPS transformation and interprets keywords such as \kvd{let!} and
\kvd{for} using special operations (monadic bind and others). The \ident{async} identifier
determines that we are writing asynchronous workflow \cite{fsharp-async} that makes it
possible to include non-blocking calls in the block.

@@

Two notable points here are that \ident{async} translated to JavaScript is restricted to a single
thread, which is not the case for ordinary F\# code (Section~\ref{sec:js-async}) and that the
\ident{HighchartOptions} object preserves some of the underlying JavaScript semantics
(Section~\ref{sec:js-lib}). Finally, the last part of the example code registers event handlers that
redraw the chart when the checkbox is clicked:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{for}~\ignp, \ignp,~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do} \\
 \quad \ident{check.click}(\kvd{fun}~\ignp~\rightarrow \ident{Async.StartImmediate}(\ident{render}()))
\end{array}
\end{equation*}
%
The \ident{click} operation (exposed by jQuery) takes a function that should be called when the
event occurs. Calling it is a side-effectful operation that registers the handler. As \ident{render}
is an asynchronous operation, we invoke it using the \ident{StartImmediate} primitive from the
F\# library, which starts the computation without waiting for the result (the only way to start
a non-blocking operation in JavaScript).

@@

The aim of this paper is not to make value judgements and argue what is better. Using the case
study as a basis, we claim that the outlined approach is just \emph{one possible} and that it
\emph{works in practice}. In the rest of the paper, we give more details about the most important
aspects of the approach and discuss alternatives.

@@

Type providers \cite{fsharp-typeprov} are a mechanism for integrating external components into a
statically typed programming language. Such components include information sources (such as World
Bank), other environments (here, JavaScript libraries via TypeScript), but type-providers can
also be used for limited meta-programming (the \ident{Asynchronous} parameter can be seen as
a form of meta-programming).

@@

\begin{itemize}
\item \emph{Type provider failure.}
  The projection (a type provider) is implemented in F\# and can fail (for example, if the internet
  is required but unavailable). However, when it succeeds, it generates valid F\# code and so
  unchecked compilation errors cannot happen.

@@

The World Bank provider \cite{fsharp-data} is a type provider for a specific data source -- it has
been designed specifically for World Bank data. This contrasts with F\# Data type providers for
working with CSV, XML and JSON data that take a sample or schema as a parameter and can thus be used
with files of any structure.

@@

\begin{itemize}
\item If you are offline when using the type provider (compiling or editing), the type provider will
  not be able to obtain the list of countries or indicators. This is an example of \emph{type provider
  failure}. In case of the World Bank, this is partly prevented by caching of the schema, but internet
  access is required when using the provider for the first time.

@@

\noindent
When contrasted with the traditional view of safety in the ML family of languages, the World Bank
type provider certainly relaxes the usual safety conditions. However, this is not the case when
compared with standard developer practices. In particular, if we wrote
$\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$ by hand, the \emph{runtime failure}
behaviour would be the same, but we would lose the useful \emph{recompilation failure}. The following
section discusses this (and other) alternatives.

@@

\vspace{-1em}
\paragraph{Accessing countries without type providers.} First, consider how we would write the code to obtain a list
of countries without using the World Bank type provider. Assuming we want only the countries and
regions in Figure~\ref{fig:wb}, we could specify the country codes and then obtain the country objects:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countryCodes}~=~\lbrack~\str{EU},~\str{CZE},~\str{GBP},~\str{USA}~\rbrack \\
 \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry})
\end{array}
\end{equation*}
%
The motivation for the case study was a data journalism application with a transparent logic
that is easy to understand (and perhaps even modify). The type provider certainly achieves these
goals better -- a country can be added just by looking through an auto-complete list and we do not
need to know its code or exact name. However, if we wanted to list all countries in a region, then
the World Bank type provider would not be helpful.

@@

\vspace{-1em}
\paragraph{Providing safer data accessors.}
The \emph{recompilation failure} when a country is removed from the dataset is an indication that
code would not behave as expected at runtime and so we find it useful. The problematic case is
a \emph{runtime failure} when a country is removed after the program is compiled. One way to
avoid this error would be to change the type of the provided members like
\ident{data.}\lident{Czech Republic} from \ident{Country} to \ident{Country~option}.

@@

Whether using option types is more desirable or not often depends on the use case (how we
want to handle errors) and reliability of the data source (countries do not disappear often).
This alternative can be supported by adding a static parameter to the type provider (like
$\ident{Asynchronous}=\kvd{true}$). For example, the CSV type provider from F\# Data has a
parameter \ident{AssumeMissingValues} that instructs the type provider to provide option values,
even if the sample does not contain them.

@@

For type providers that provide access to external information sources, we can formulate a
\emph{relativized} notion of type safety. A full formalization is beyond the scope of this
paper, but the following provides an outline of such a theorem.

@@

JavaScript has become the \emph{lingua franca} of the web and an increasing number of programming
languages provide a way of compiling to JavaScript. In F\#, the first project was F\# WebTools
\cite{fsharp-webtools} in 2007. A more recent and complex framework called Websharper \cite{websharper-guis}
is available with full commercial support. In this paper, we use FunScript which is a lightweight
library focused just on translating F\# to JavaScript.

In this section, we discuss the options available when targeting JavaScript. The compilation to
JavaScript can be implemented as compiler back-end or as a library based on meta-programming
(Section~\ref{sec:js-meta}), there are different approaches to base libraries (Section~\ref{sec:js-lib}),
asynchronous computing (Section~\ref{sec:js-async}) and the semantic mismatch between the host
language and JavaScript (Section~\ref{sec:js-relative}).

@@

In FunScript, we typically need to translate the entire source file. To allow this, we instruct the
compiler to store code as data for the whole module using the \ident{ReflectedDefinition} attribute.
This is done at the beginning of the file (which we omitted in Section~\ref{sec:case}):
%
\begin{equation*}
\begin{array}{l}
 \lbrack\langl\ident{ReflectedDefinition}\rangl\rbrack \\
 \kvd{module}~\ident{Program}
 \\[0.5em]
 \kvd{open}~\ident{FunScript} \\
 \kvd{open}~\ident{FunScript.TypeScript} \\
 \kvd{open}~\ident{FSharp.Data}
\end{array}
\end{equation*}
%
When marked with \ident{ReflectedDefinition}, the F\# compiler stores the body of all functions and
methods in the marked module as \emph{quotations} that can be retrieved at run-time. Note that this does
not store the AST for the entire file -- for example, information about declared types needs to be
obtained separately using the .NET reflection mechanism.

@@

Using the lightweight meta-programming approach in FunScript would not be appropriate if we wanted to
compile arbitrary existing F\# source code to JavaScript without any modifications. For that, using a
compiler back-end is a better choice. However, for the task solved in our case study (compiling
newly written F\# code), the approach works well. A related question is providing mappings for
standard (in our case, F\# and .NET) libraries used by the code. We discuss this in the next section.

@@

The F\# code in our case study uses a number of libraries. This includes the F\# core library (for
example, the \ident{List.map} function), standard .NET libraries (iteration using the \kvd{for} loop
uses .NET \ident{IEnumerable} interface) and JavaScript libraries. Accessing these libraries in
FunScript follows a similar approach to F\# WebTools \cite{fsharp-webtools}.

@@

FunScript implements (explicit) mappings for standard F\# and .NET libraries that are useful in
JavaScript (collections, string manipulation, date and other). Those are either reimplemented in
F\# with the \ident{ReflectedDefinition} attribute, or mapped to standard JavaScript library.

@@

\noindent
In summary, the approach used by F\# is to reuse as much of F\# and .NET as possible. This
makes writing and reusing code easier and lets developers use existing familiar libraries and
idioms. However, it means that the semantics of F\# running as JavaScript is not the same as the
semantics of F\# running as compiled code. We return to this topic in Section~\ref{sec:js-relative}.
Before that, the following section gives one more example where similar approach is used.

@@

In F\#, asynchronous workflows \cite{fsharp-async} serve two purposes. First, they provide a
way for running multiple tasks in parallel. Second, they make it possible to write long-running
non-blocking code without the use of explicit callbacks. In our case study, we only use the
latter aspect. When loading data for selected countries in a loop, we fetch data asynchronously
(using an AJAX call). Under the cover, JavaScript triggers a callback, which then resumes the
asynchronous workflow.

@@

In ordinary F\#, asynchronous workflows can be started in a number of ways: \ident{Async.RunSynchronously}
starts the work and blocks until it completes; \ident{Async.Start} starts the work in the background
and \ident{Async.\-Start\-Immediate} starts the work on the current thread and runs callbacks
in the same synchronization context (essentially on the same thread). However, only the last one
can be mapped to JavaScript (FunScript translator will fail if the other two are used).

@@

There are a number of projects that solve similar problems to the ones described in this paper.
In the F\# ecosystem, the first project compiling to JavaScript was F\# WebTools \cite{fsharp-webtools},
which also supported asynchronous computations (although using a separate computation type)
and interop with JavaScript. More recently, WebSharper \cite{websharper-piglets,websharper-guis}
is a more complete framework for web development that also provides composable abstractions for
building forms (based on formlets \cite{links-formlets}) and entire applications.

@@

The key idea of this paper is that writing modern applications for the web requires us to
reconsider many assumptions that functional language designers take for granted. The novelty
of this paper is not in any single technology it presents, but in the combination it shows.
For this reason, we used the form of a case study, using a significant example to guide our
discussion.
